---
title: "Protist communities along a surface-groundwater gradient in the Danube wetlands of Vienna" 
author:  "Angela Cukusic"
output:
  html_document:
    toc: true               # Table of contents
    toc_float: true         # Keep the table of contents floating on the page
    code_download: true     # Enable code download button
    code_copy: true         # Enable code copy button
    number_sections: true   # Numbering the sections
    theme: cerulean           # Choose a Bootstrap theme (check available themes)
    highlight: tango        # Syntax highlighting style 
---

```{r setup, include=FALSE}
# For width of code chunks and scroll bar 
options(width=250)

knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      cache = TRUE,
                      include = TRUE,
                      warning = FALSE,
                      collapse = FALSE,
                      message = FALSE,
                      engine = "R", 
                      # Chunks will always have R code, unless noted
                      error = TRUE)

```

# load packages
```{r}
library(tidyverse)
library(stringr)
library(decontam)
library(readxl)
library(vegan)
library(patchwork)
library(ggpmisc)
library(MASS)
library(microViz)
library(microbiome)
library(car)
library(corrplot)
library(magrittr)
library(compositions)
library(microeco)
library(cowplot)

library(DESeq2)


pal <- c("gray90", "gray20")
```


# load data
```{r}
# asv table
asv_table_SSU <-  read.csv("~/faks/poster_fungi/poster_fungi_jmf_2023-05-16/JMF-2301-09__all__rRNA_SSU_Euk_D__JMFR_MSHI_KP237/DADA2_counts_as_matrix.tsv", row.names=1,  check.names = FALSE, sep = "\t")
names(asv_table_SSU)<- sub("C","", names(asv_table_SSU))
# the first 78 samples are rna, so we take the second half
asv_table <- asv_table_SSU[, -c(1:78)]


# tax table
tax_table_SSU_RP2 <- read.delim("~/faks/poster_fungi/poster_fungi_jmf_2023-05-16/JMF-2301-09__all__rRNA_SSU_Euk_D__JMFR_MSHI_KP237/DADA2_ASVs.rRNA_SSU.PR2_reference.DADA2_classified.tsv", row.names=1)

#env table
env_data <- read_excel("~/faks/christina/data/data_angela/env_combined.xlsx") 
```



# remove the blanks
```{r}
jmf_blank_info <- read_excel("~/faks/christina/data/data_angela/jmf_2301_09_esw_d15.xlsx") 


jmf_blank_info <- 
# choose only dna  
  jmf_blank_info[ -c(1:78),] %>% 
  dplyr::select("JMF sample ID", "Sample description", "Date") %>% 
  rename("Sample_desc" = "Sample description",
          "JMF.ID" = "JMF sample ID") %>% 
# add blank or sample column  
  mutate(blank_sample = case_when(startsWith(Sample_desc, "Blank") ~ "blank",
                                             TRUE ~ "sample")) %>% 
  dplyr::filter(blank_sample == "sample")
 
 
# remove blanks from the asv table
asv_table <- asv_table[, colnames(asv_table) %in% jmf_blank_info$JMF.ID]
```




# connect sequincing id to sample_id
```{r}

# find the unique sample name first based on well id and sampling date
jmf_blank_info <- jmf_blank_info  %>% 
  mutate (well_id = case_when(startsWith(Sample_desc, "ESW") ~ "ESW_", 
                              startsWith(Sample_desc, "D05") ~ "D05p_", 
                              startsWith(Sample_desc, "D10") ~ "D10p_", 
                              startsWith(Sample_desc, "D15") ~ "D15p_"   )) %>% 
   mutate(sample_id = paste(well_id, Date)) %>% 
   dplyr::filter(well_id %in% c("ESW_", "D15p_"))
jmf_blank_info$sample_id <- gsub(" ", "", jmf_blank_info$sample_id)
jmf_blank_info$sample_id <- gsub("-", "", jmf_blank_info$sample_id)

# clean the Unbekannt one (JMF-2301-09-0109)that we do not know the id of
jmf_blank_info <- jmf_blank_info %>% dplyr::filter(Sample_desc != "Unbekannt")


env_lobau_final <- inner_join(jmf_blank_info, env_data, by = "sample_id")
rownames(env_lobau_final) <- env_lobau_final$JMF.ID

# rename the asv_table also
asv_table %>% 
  rownames_to_column("taxa") %>% 
  pivot_longer(cols = - taxa, names_to = "JMF.ID", values_to = "counts" ) %>% 
  left_join(., jmf_blank_info[,c("sample_id", "JMF.ID")], by = "JMF.ID") %>% 
  dplyr::select(!"JMF.ID") %>% 
# if there is replicates for the same sample_id add it up  
  group_by(taxa, sample_id) %>%
  summarise(counts = sum(counts)) %>%
  ungroup() %>% 
  pivot_wider(names_from = "sample_id", values_from = "counts") %>%
  column_to_rownames(var="taxa") -> asv_table


# the last column is NA -  the Unbekannt one?
# cannot be sure so just remove it
asv_table <- asv_table[, 1:46]
```


# clean up env data
```{r}
# since we removed the replicates from asv_table remove them from env_table

env_lobau_final %>% dplyr::filter(! JMF.ID %in% c("JMF-2301-09-0152", 
                                                   "JMF-2301-09-0147", 
                                                   "JMF-2301-09-0133",
                                                   "JMF-2301-09-0082", 
                                                   "JMF-2301-09-0083",
                                                   "JMF-2301-09-0084", 
                                                   "JMF-2301-09-0102", 
                                                   "JMF-2301-09-0106", 
                                                   "JMF-2301-09-0104", 
                                                   "JMF-2301-09-0139",
                                                   "JMF-2301-09-0100", 
                                                   "JMF-2301-09-0092") ) -> env_lobau_final

# since i know which ones are replicates it is easiest for me to just remove those jmfs



# add season category
env_lobau_final <- env_lobau_final %>% 
mutate(season = case_when ( 
     grepl('2021-05|2021-04|2022-04|2021-03|2022-03|2022-03|2022-05', Date) ~ "spring",
     grepl('2020-07|2022-08|2021-06|2022-06|2021-07|2022-08', Date)  ~ "summer",
     grepl('2020-09|2020-10|2022-10|2021-10|2022-11|2020-11|2021-11', Date) ~ "fall",
     grepl('2020-12|2021-12|2021-01|2022-01|2021-02|2022-02', Date) ~ "winter"   ))
```

We are left with 46 samples.


# clean up tax data
For now just remove the Metazoa, Fungi and unclassified divisions.
```{r}
tax_table <- tax_table_SSU_RP2 %>% 
  dplyr::filter(!Supergroup == "") %>% 
  dplyr::filter(!Division %in% c("Fungi", "Metazoa", "") )

# remove the taxa also from asv_table
asv_table <- asv_table[which(rownames(asv_table) %in% rownames(tax_table)) ,]
```




# Read numbers
```{r}
read_data <- asv_table %>% 
colSums() %>% # how many reads per sample (find sums per each column)
as.data.frame() %>% 
rownames_to_column("sample_id") %>% 
rename("reads" = ".") 

summary(read_data)
sd(read_data$reads)

boxplot(read_data$reads)


# mean: 17356  +- 10557.83



```

# Count numbers
Search for the count numbers - considering the elevated numbers of contaminated aquifers in comparison to pristine aquifers, it could be hypothesized there is a bigger count number in the surface water?
```{r}

# sample sizes
table(env_lobau_final$season, env_lobau_final$well_id.y)
# fairly similar sample sizes



asv_table %>% 
  rownames_to_column("taxa") %>% 
  pivot_longer(cols = -taxa, values_to = "counts", names_to = "sample_id") %>% 
  left_join(., env_lobau_final[, c("sample_id", "well_id.y", "season")], by = "sample_id") %>% 
  
# do not count by individual taxa, look at per sample count  
  group_by(sample_id) %>% 
  mutate(counts = sum(counts)) %>% 
  ungroup() %>% 
  
  ggplot(aes(x= well_id.y, y=counts, fill = well_id.y)) +
  geom_boxplot(alpha = 0.7) + 
  scale_y_log10(labels =  function(x) format(x, scientific = TRUE)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "gray"),
        legend.position =c(0.1, 0.9),
        axis.text.x = element_text(face = "bold", size = 12), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14),
        legend.text = element_text(size = 11) ) +
  scale_fill_manual(values = pal) +
  labs(x = "Well ID", y = "Total 18S counts of sample",
       fill = "Well ID")   -> p_counts2


```


```{r}
asv_table %>% 
  rownames_to_column("taxa") %>% 
  pivot_longer(cols = -taxa, 
               values_to = "counts", 
               names_to = "sample_id") %>% 
  left_join(., env_lobau_final[, c("sample_id", "well_id.y", "season")], 
            by = "sample_id") %>% 
  
# do not count by individual taxa, look at per sample count  
  group_by(sample_id, well_id.y) %>% 
  summarise(counts = sum(counts)) %>% 
  ungroup() -> data3



data3[data3$well_id.y == "ESW","counts"] %>% as.data.frame -> counts_esw
data3[data3$well_id.y == "D15","counts"] %>% as.data.frame -> counts_D15

# Shapiro-Wilk test
shapiro.test(counts_esw$counts)
shapiro.test(counts_D15$counts)

# Levene's test
leveneTest(data3$counts ~ well_id.y, data=data3)

# not normal - use non parametric

wilcox.test(counts_esw$counts, counts_D15$counts)
  
```


# by season
```{r}

asv_table %>% 
  rownames_to_column("taxa") %>% 
  pivot_longer(cols = -taxa, values_to = "counts", names_to = "sample_id") %>% 
  left_join(., env_lobau_final[, c("sample_id", "well_id.y", "season")], by = "sample_id") %>% 
  
# do not count by individual taxa, look at per sample count  
  group_by(sample_id) %>% 
  mutate(counts = sum(counts)) %>% 
  ungroup() %>% 
  
  ggplot(aes(x= season, y=counts, fill = well_id.y)) +
  geom_boxplot(alpha = 0.7) + 
  scale_y_log10(labels =  function(x) format(x, scientific = TRUE)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "gray"),
        legend.position =c(0.1, 0.9),
        axis.text.x = element_text(face = "bold", size = 12), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14),
        legend.text = element_text(size = 11) ) +
  scale_fill_manual(values = pal) +
  labs(x = "Season", y = "Total 18S counts of sample",
       fill = "Well ID") +
  scale_x_discrete(limits = c("fall", "winter", "spring", "summer"))   -> p_counts
p_counts
```



# Rarefaction
```{r}

rrarefy(t(asv_table), sample = 4000) %>% t() %>% as.data.frame() -> asv_table_r

asv_table_r %>% colSums() %>%  plot()

# there is 6 sample that have less than 4000 counts
# remove them from asv_table
asv_table_r[,colSums(asv_table_r) == 4000] -> asv_table_r
# remove them from env data
env_lobau_final[env_lobau_final$sample_id %in%  colnames(asv_table_r),] -> env_lobau_final_r
```



# ordering
```{r}
env_lobau_final_r[order(env_lobau_final_r$sample_id), ] -> env_lobau_final_r
rownames(env_lobau_final_r) <- env_lobau_final_r$sample_id
identical(rownames(env_lobau_final_r), colnames(asv_table_r))
```


# Diversity
```{r}

shannondiv <- vegan::diversity(t(asv_table_r))

as.data.frame(shannondiv) %>% 
  rownames_to_column("sample_id") %>% 
  left_join(., env_lobau_final_r, by = "sample_id") %>% 
# plot
  ggplot(aes(x= season, y=shannondiv, fill = well_id.y)) +
  geom_boxplot(alpha = 0.7) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "gray"),
        legend.position = c(0.1, 0.1),
        axis.text.x = element_text(face = "bold", size = 12), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14)) +
  scale_fill_manual(values = pal) +
  labs(x = "Season", y = NULL, fill = "Well ID")  +
  scale_x_discrete(limits = c("fall", "winter", "spring", "summer")) -> p_div

p_div


```

```{r}

as.data.frame(shannondiv) %>% 
  rownames_to_column("sample_id") %>% 
  left_join(., env_lobau_final_r, by = "sample_id") -> data3



# fall
wilcox.test(data3 %>% dplyr::filter(season == "fall" & well_id.y == "D15") %>% .$shannondiv, 
           data3 %>% dplyr::filter(season == "fall" & well_id.y == "ESW")  %>% .$shannondiv )

# spring
wilcox.test(data3 %>% dplyr::filter(season == "spring" & well_id.y == "D15") %>% .$shannondiv, 
           data3 %>% dplyr::filter(season == "spring" & well_id.y == "ESW")  %>% .$shannondiv )

# summer
wilcox.test(data3 %>% dplyr::filter(season == "summer" & well_id.y == "D15") %>% .$shannondiv, 
           data3 %>% dplyr::filter(season == "summer" & well_id.y == "ESW")  %>% .$shannondiv )


```


```{r}

as.data.frame(shannondiv) %>% 
  rownames_to_column("sample_id") %>% 
  left_join(., env_lobau_final_r, by = "sample_id") %>% 
# plot
  ggplot(aes(x= well_id.y, y=shannondiv, fill = well_id.y)) +
  geom_boxplot(alpha = 0.7) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "gray"),
        legend.position ="none",
        axis.text.x = element_text(face = "bold", size = 12), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14)) +
  scale_fill_manual(values = pal) +
  labs(x = "Well ID", y = "Shannon's diversity index")  -> p_div2

p_div2

```

```{r}

data3[data3$well_id.y == "ESW","shannondiv"] %>% as.data.frame -> ens_esw
data3[data3$well_id.y == "D15","shannondiv"] %>% as.data.frame -> ens_D15

# Shapiro-Wilk test
shapiro.test(ens_esw$.) # normal
shapiro.test(ens_D15$.) # not normal

# Levene's test
leveneTest(data3$shannondiv ~ well_id.y, data=data3)

# not normal - use non parametric
wilcox.test(ens_esw$., ens_D15$.)
```


```{r}

p_div2 + p_div  + plot_layout(widths = c(1,2))
```

# nmds
```{r}

# do an ordination with a hellinger transformation 
sp.dist <-  vegdist(decostand( t(asv_table_r),
                               method = "hellinger") )



nmds <- metaMDS(sp.dist, k = 2, autotransform = FALSE)
plot(nmds)


as.data.frame(nmds$points) %>% 
  cbind(., env_lobau_final_r) %>% 
  
  ggplot(aes(x=MDS1,y=MDS2) ) +
  geom_point(aes( fill= well_id.y, shape = season), 
             size = 4, alpha = 0.6, stroke = 0.7) +  
# coloring  
  #scale_color_manual(values=pal) +
  #scale_fill_manual(values=pal) +
# add stress value
  annotate(geom="text",x=0.7, y=1.2, 
           label="stress =", color="black") +   
  annotate(geom="text", x=1.1, y=1.2, 
           label=round(nmds$stress,4), color="black") +
# ellipse  
  stat_ellipse(aes(colour = well_id.y), linewidth = 0.9) + 
  theme_bw() +
# visual parameters 
  theme(panel.grid = element_blank(),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "gray")) +
  labs(color = "Well ID", shape = "Season") +
  guides(fill = "none") + 
  scale_shape_manual(values = c(21, 22, 23, 24)) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) 

```



```{r}

# well id
# first test assumption: equal dispersion
disp_aq <- betadisper( 
  vegdist(t(asv_table_r),
          method = "bray"), 
          env_lobau_final_r$well_id.y ) 

#p>0.05 is what we need - homogenous variances
anova(disp_aq) #no sign. differences in dispersion


# now test actual ordination
permanova_aq <- adonis2(t(asv_table_r) ~ 
          env_lobau_final_r$well_id.y , 
                         method = "bray" , 
                        permutations = 999)
permanova_aq # not sign


# season
# first test assumption: equal dispersion
disp_aq <- betadisper( 
  vegdist(t(asv_table_r),
          method = "bray"), 
          env_lobau_final_r$season ) 

#p>0.05 is what we need - homogenous variances
anova(disp_aq) #no sign. differences in dispersion


# now test actual ordination
permanova_aq <- adonis2(t(asv_table_r) ~ 
          env_lobau_final_r$season , 
                         method = "bray" , 
                        permutations = 999)
permanova_aq # not sign





adonis2(t(asv_table_r) ~ 
          env_lobau_final_r$season *  env_lobau_final_r$well_id.y, 
                         method = "bray" , 
                        permutations = 999)

```


# env variables


```{r}
my_labeller2 <- as_labeller(c(
  ec_uS_cm = "EC ~ (µS/cm)",
  pH = "pH",
  oxygen_mg_L="O[2] ~ (mg/L)",
  doc_mg_L = "DOC ~ (mg/L)",
  dic_mg_L = "DIC ~ (mg/L)", 
  co2_uM = "CO[2] ~ (uM)"),
  default = label_parsed)

my_labeller1 <- as_labeller(c(
 atp_internal_pM = "ATP ~  (pM/L)",
  cells_per_mL = "TCC ~ (cells/mL)"),
  default = label_parsed)


        

```


```{r}
env_lobau_final_r %>% 
  dplyr::select("well_id.y", "pH", "ec_uS_cm", "oxygen_mg_L", "doc_mg_L", "dic_mg_L", 
                 "co2_uM") %>% 
  dplyr::filter(doc_mg_L < 30) %>% 
dplyr::filter(co2_uM < 1000) %>% 
pivot_longer(cols = -well_id.y, values_to = "values", names_to = "names") %>% 
  ggplot(aes(x= well_id.y, y= values, fill = well_id.y)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~names, scales = "free_y", ncol = 2,  
             strip.position = "left", 
             labeller = my_labeller2) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
         axis.text.x = element_text(face = "bold", size = 12), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14),
        plot.margin = unit(c(0, 0.5, 0, 0), "null"), 
        strip.text = element_text(size = 12)) +
  scale_fill_manual(values = pal) + 
  labs(y = NULL, x = "Well ID") -> plot_rest


env_lobau_final_r %>% 
  dplyr::select("well_id.y", "atp_internal_pM",
                "cells_per_mL") %>% 
pivot_longer(cols = -well_id.y, values_to = "values", names_to = "names") %>% 
  ggplot(aes(x= well_id.y, y= values, fill = well_id.y)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10() +
   facet_wrap(~names, scales = "free_y", ncol = 2,  
              strip.position = "left", 
             labeller = my_labeller1) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
         axis.text.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14),
        axis.ticks.x = element_blank(),
        strip.text = element_text(size = 12)) +
  scale_fill_manual(values = pal) + 
  labs(y = NULL, x = NULL) -> plot_atp

plot_atp + plot_rest + plot_layout(heights = c(1,3))

```
# composition
```{r}
tax_table %>% 
 mutate(Class = ifelse(Class == "", paste0("Unclassified ", Division), Class)) -> tax_table

asv_table_r %>% 
  mutate_if(is.integer, as.numeric) %>% 
  mutate(class = tax_table$Class) %>%
  pivot_longer(cols = -class, 
               values_to = "abundance", 
               names_to = "sample") %>%
  left_join(.,env_lobau_final_r, by = c("sample" = "sample_id")) %>%  
  dplyr::select ("abundance", "class", "well_id.y") %>%
  mutate(well_id.y = as.factor(well_id.y),
         class = as.factor(class)) %>%  
# to find relative abundance within each category  
  dplyr::group_by(well_id.y) %>%
  mutate (rel_abund = abundance / sum(abundance)) %>%
  ungroup() %>%
# to sum up all the counts of one class from all the samples  
  dplyr::group_by(well_id.y, class) %>%
  summarise(rel_abund = sum(rel_abund)) %>%
  ungroup() %>%
# those very rare categorize as Diverse others 
  dplyr::group_by(well_id.y, class) %>%
  mutate(class = case_when(rel_abund< 0.015~ "Diverse others", 
                           TRUE ~ class)) %>%
  ungroup() %>%
# final regrouping after categorizing for Diverse others  
  dplyr::group_by(well_id.y, class) %>%
  summarise(rel_abund = sum(rel_abund)) %>%
  ungroup() %>%

# plot
  ggplot(aes(x = well_id.y, y = rel_abund, fill = class)) + 
  geom_bar(stat = "identity", color = "black",
           ) +
  #scale_fill_manual(values=c25) +
  theme(panel.grid = element_blank(), legend.position="right",
        legend.text = element_text(size = 6),      # Adjust the text size
        legend.title = element_text(size = 6),
        legend.key.size = unit(0.4, "cm"),
        axis.title = element_text(size=14,face="bold"),
        axis.text.y = element_text(size=10,face="bold"),
        axis.text.x = element_text(angle = 90, face = "bold")) + 
  labs(x = "Temperature categories",
       y = "Relative abundance", 
       fill = "Class") +
  guides(fill=guide_legend(ncol = 1)) 


```
## by sample site
```{r}

library(phyloseq)
library(microshades)

as.data.frame(env_lobau_final_r) %>% column_to_rownames("sample_id") -> env.data

ps_dna <- phyloseq(otu_table(as.matrix(asv_table_r), taxa_are_rows=TRUE), 
                   tax_table(as.matrix(tax_table) ) , 
                   sample_data(env.data ) )



 mdf_prep <- ps_dna %>%
         tax_glom("Class") %>%  
         merge_samples( group = "well_id.y") %>%
         phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
         psmelt() %>%
         filter(Abundance > 0) %>% 
         dplyr::select("Abundance", "Sample", "Division", "Class") %>%
         mutate(Division = as.factor(Division),
                Class = as.factor(Class))
 
# find the 4 most abundant division 
 mdf_prep %>% group_by(Division) %>% 
   summarise(Abundance = sum (Abundance)) %>% 
   slice_max(order_by =Abundance, n= 4) %>% 
   as.data.frame() %>% 
   dplyr::select("Division") %>% unique()
 # c( Ochrophyta, Ciliophora, Discoba, Conosa, Dinoflagellata,	Cryptophyta, Cercozoa )
 
 
 

color_objs_GP <- create_color_dfs(mdf_prep,selected_groups = c("Ochrophyta", 
                                                               "Ciliophora", 
                                                               "Discoba",
                                                               "Conosa", 
                                                               #"Dinoflagellata",
                                                               #"Cryptophyta",
                                                               "Cercozoa"
                                                               ) , cvd = TRUE, 
                                  group_level = "Division", subgroup_level = "Class",
                                  top_n_subgroups = 4)

# Extract
mdf_GP <- color_objs_GP$mdf
cdf_GP <- color_objs_GP$cdf


plot <- plot_microshades(mdf_GP, cdf_GP) +
  scale_y_continuous(labels = scales::percent, 
                     expand = expansion(0)) +
  theme_bw() +
  theme(legend.position = "none",
        panel.grid = element_blank(),
         axis.text.x = element_text(face = "bold", size = 12), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14)) 
plot


GP_legend <- custom_legend(mdf_GP, 
                          cdf_GP, 
                          group_level = "Division", 
                          subgroup_level = "Class",
                          legend_key_size = 1,
                          legend_text_size = 12)
cowplot::plot_grid(plot, GP_legend,  rel_widths = c(1, .5))


```


## by season
```{r}
mdf_prep_s <- ps_dna %>% tax_glom("Class") %>% 
  psmelt() %>% 
  group_by(season, well_id.y, Class) %>% 
  mutate(Abundance = sum (Abundance)) %>% 
  ungroup()  %>% 
  dplyr::group_by(season, well_id.y) %>%
  mutate (rel_abund = Abundance / sum(Abundance)) %>%
  ungroup() %>% 
  dplyr::select( "rel_abund", "well_id.y", "Division", "Class" , "season") %>% 
  dplyr::rename("Sample" = "well_id.y",
                "Abundance" = "rel_abund") %>% 
  as.data.frame()



# find the 4 most abundant division 
 mdf_prep_s %>% group_by(Division) %>% 
   summarise(Abundance = sum (Abundance)) %>% 
   slice_max(order_by = Abundance, n = 4) %>% 
   as.data.frame() %>% 
   dplyr::select("Division") %>% unique()
 # c( Ochrophyta, Ciliophora, Discoba, Conosa, Dinoflagellata,	Cryptophyta, Cercozoa )
 
 
 

color_objs_GP <- create_color_dfs(mdf_prep_s, 
                                  selected_groups = c("Ochrophyta", 
                                                      "Ciliophora", 
                                                      "Discoba",
                                                      "Conosa", 
                                                      #"Dinoflagellata",
                                                      #"Cryptophyta",
                                                      "Cercozoa" ), 
                                  cvd = TRUE, 
                                  group_level = "Division", 
                                  subgroup_level = "Class",
                                  top_n_subgroups = 4)

# Extract
mdf_GP <- color_objs_GP$mdf 
# dataset with      Abundance Sample  Division   
cdf_GP <- color_objs_GP$cdf



plot <- plot_microshades(mdf_GP, cdf_GP) +
  scale_y_continuous(labels = scales::percent, 
                     expand = expansion(0)) +
  theme_bw() +
  facet_grid(~season) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
         axis.text.x = element_text(face = "bold", size = 12), 
        axis.text.y = element_text(face = "bold", size = 11), 
        axis.title = element_text(face = "bold", size = 14)) 


GP_legend <- custom_legend(mdf_GP, 
                          cdf_GP, 
                          group_level = "Division", 
                          subgroup_level = "Class",
                          legend_key_size = 1,
                          legend_text_size = 12)
cowplot::plot_grid(plot, GP_legend,  rel_widths = c(1, .5))
```

```{r}

asv_table_r %>% 
  mutate_if(is.integer, as.numeric) %>% 
  mutate(Division = tax_table$Division) %>%
  pivot_longer(cols = -Division, 
               values_to = "abundance", 
               names_to = "sample") %>%
  mutate(Division = as.factor(Division) ,
         sample = as.factor(sample))%>% 
  
  group_by(sample, Division) %>% 
  summarise(abundance = sum(abundance)) %>% 
  ungroup() %>% 
  
  left_join(.,env_lobau_final_r, by = c("sample" = "sample_id")) %>%  
  dplyr::select ("abundance", "Division", "well_id.y", "sample") %>% 
  mutate(well_id.y = as.factor(well_id.y)) %>%  
# to find relative abundance within each sampling site, but also within each sample  
  dplyr::group_by(sample) %>%
  mutate (rel_abund = abundance / sum(abundance)) %>%
  ungroup() %>%
  dplyr::filter(Division %in% c("Ochrophyta",
                                "Ciliophora",
                                "Dinoflagellata", 
                                "Cryptophyta",
                                "Conosa", 
                                "Cercozoa",
                                "Discoba")) %>% 
  mutate(Ecology  = case_when(Division %in% c( "Ciliophora", 
                                               "Cercozoa") ~ "consumer", 
                              Division %in% c("Discoba", "Dinoflagellata") ~  "myxothrop", 
                              Division %in% c("Conosa") ~  "parasite", 
                              Division %in% c("Ochrophyta", "Cryptophyta") ~ "phototroph")) %>% 
  dplyr::filter(rel_abund >0 ) %>% 

  ggplot(aes(y=reorder(Division, rel_abund), x=rel_abund)) + 
  geom_point(aes(color = Ecology, fill = Ecology), alpha =0.2, size = 3) +
  geom_point(aes(color = Ecology, fill = Ecology), alpha =0.2, size = 2) +
  geom_boxplot(aes(color = Ecology, fill = Ecology), alpha =0.3, outlier.colour = NA) +
  facet_wrap(~well_id.y, ncol = 1) + 
  theme_bw() +
  labs(x  = "Relative abundance %", y= NULL) +
  scale_fill_manual(values = c("blue", "orange2", "gold", "darkgreen")) + 
  scale_color_manual(values = c("blue", "orange2", "gold", "darkgreen")) +
  theme( legend.background = element_blank(),
        legend.box.background = element_rect(colour = "gray"),
        legend.position = c(0.8, 0.2), 
        panel.grid = element_blank(), 
        strip.text = element_text(size =10)) 

```

```{r}
asv_table_r %>% 
  mutate_if(is.integer, as.numeric) %>% 
  mutate(Division = tax_table$Division) %>%
  pivot_longer(cols = -Division, 
               values_to = "abundance", 
               names_to = "sample") %>%
  mutate(Division = as.factor(Division) ,
         sample = as.factor(sample))%>% 
  
  group_by(sample, Division) %>% 
  summarise(abundance = sum(abundance)) %>% 
  ungroup() %>% 
  
  left_join(.,env_lobau_final_r, by = c("sample" = "sample_id")) %>%  
  dplyr::select ("abundance", "Division", "well_id.y", "sample") %>% 
  mutate(well_id.y = as.factor(well_id.y)) %>%  
# to find relative abundance within each sampling site, but also within each sample  
  dplyr::group_by(sample) %>%
  mutate (rel_abund = abundance / sum(abundance)) %>%
  ungroup() %>%
  dplyr::filter(Division %in% c("Ochrophyta",
                                "Ciliophora",
                                "Dinoflagellata", 
                                "Cryptophyta",
                                "Conosa", 
                                "Cercozoa",
                                "Discoba")) %>% 
  
   dplyr::group_by(well_id.y, Division) %>%
  summarise(mean = mean(rel_abund) *100, sd = sd(rel_abund)*100) %>%
  ungroup() %>% 
  arrange(mean)
```

# indicator species
```{r}
library(indicspecies)
# community matrix as df with sites in rows
t(asv_table_r) %>% as.data.frame() -> sp.df
# make a vector with sampling sites
sp.df %>% rownames_to_column("sample_name") %>% 
  dplyr::select("sample_name") %>% 
  left_join(.,env_lobau_final_r, by = c("sample_name" = "sample_id")) %>% 
  dplyr::select("well_id.y") %>% 
  table()

 c(rep(1, 17), rep(2, 23) ) -> group_vect


indval <- multipatt(sp.df, 
                    group_vect,
                    func="IndVal.g",  
                    duleg=TRUE,  
                    control = how(nperm=999)) 
indval$sign
# the NA p values are the species present in both sites





# adjust with fdr:

# see how many are significant before the fdr
indval[["sign"]][["p.value"]] %>% as.data.frame() %>% 
  filter(.<0.05) # 908

# after fdr
p.adjust(indval[["sign"]][["p.value"]], method="BH") %>% 
  as.data.frame() %>% 
  filter(.<0.05) # 597


# add species names 
p.adjust(indval[["sign"]][["p.value"]], method="BH") %>% 
  as.data.frame() %>% 
  dplyr::rename("adj.p" = ".") %>% 
# after correcting add the names from the original output  
  cbind(indval$sign) %>% 
# filter now only the truly sign ones
  dplyr::filter(adj.p < 0.05) %>% 
  dplyr::select("adj.p", "s.1", "s.2", "stat") %>% 
  rownames_to_column("ASVs") %>% 
  left_join(., tax_table %>% rownames_to_column("ASVs"), 
            by = "ASVs") -> species.df

```




# plot df

## bubble plot
I remove the adj p because I dont use it for plotting.
```{r}
species.df %>% 
  mutate (Sample_site = ifelse(s.1 == 1, "D15", "ESW") ) %>% 
  dplyr::select("Division", "Family", "Sample_site", "stat") %>%
  dplyr::filter(Family != "") %>% 
  arrange(desc(stat) ) %>% 
  slice_max(stat, n= 70) %>% 
  ggplot() +
  geom_point(aes(x = as.factor(Sample_site), y= Family, 
                 color = Division, cex = stat )
             #shape = 21, color = "black"
             ) + 
  theme_bw() + 
  scale_color_manual(values = 
#distinct_palette(n = NA, pal = "brewerPlus") 
  c("#A6CEE3",  "#1F78B4",  "#B2DF8A", "#33A02C",  "#E31A1C",   "#FDBF6F",
    "#CAB2D6",   "#6A3D9A",  "#1ff8ff",  "#83d5af" ,  "#E7298A",
    "#66A61E",  "#A6761D",  "#4b6a53",   "#b249d5",    "#7edc45",
    "#5c47b8",   "#cfd251",   "#ff69b4", "#69c86c")) -> plot_ind

plot_ind

 #"#cd3e50"    "#da6130"   "#5e79b2"   "#c29545"   "#532a5a"   "#5f7b35"  
 #"#c497cf"   "#773a27"   "#7cb9cb"   "#594e50"   "#d3c4a8"   "#c17e7f" ))
  




```


## lollipop plot
```{r}
species.df %>% 
  mutate (Sample_site = ifelse(s.1 == 1, "D15", "ESW") ) %>% 
  dplyr::select("Division", "Family", "Sample_site", "stat") %>%
  dplyr::filter(Family != "") %>% 
  arrange(desc(stat) ) %>% 
  slice_max(stat, n= 70) %>% 
  ggplot() +
  geom_point(aes(x = stat, y= Family, 
                 color = Sample_site),
             #shape = 21, color = "black"
             size = 3 ) + 
  geom_segment(aes(x=0.8, xend= stat, y=Family, yend =Family)) +
  theme_bw()
```

## try to cut sp down more 

This function allows reducing drastically the number of species combinations to be retained for a given target site group.
```{r}
sc <- indicators(X=sp.df, cluster=group_vect, 
                 group=2, 
                 max.order = 1, 
                 verbose=TRUE, 
                 At=0.5, Bt=0.2)


pruneindicators(sc, At = 0.2,
                Bt = 0.3,
                sqrtIVt = 0,
                alpha = 1,
                max.indicators = 4,
                verbose = FALSE)
```




# differential abundance

Highly disputable since I needed to add 1 to each value in asv table in order to be able to compute geometric means...
```{r}
# for daa we need unrarified data

# remove those with few reads
asv_table[,colSums(asv_table) > 4000] -> asv_table_unrar
# remove them from env data
env_lobau_final[env_lobau_final$sample_id %in%  colnames(asv_table),] -> env_lobau_final_unrar



as.data.frame(env_lobau_final_unrar) %>% column_to_rownames("sample_id") -> env.data_unrar

ps_dna_unrar <- phyloseq(otu_table(
                                  ( as.matrix(asv_table_unrar) + 1) , 
                                   taxa_are_rows=TRUE), 
                   tax_table(as.matrix(tax_table) ) , 
                   sample_data(env.data_unrar ) )


# make a deseq object
dds.data = phyloseq_to_deseq2( ps_dna_unrar, ~ well_id.y)


dds = DESeq(dds.data)
dds$well_id.y
res <- results(dds)  

res <- res[order(res$padj, na.last=NA), ]
sigtab <- res[(res$padj < alpha), ] %>%  as.data.frame() %>% 
  rownames_to_column("ASV") %>% 
  left_join(. , tax_table %>% rownames_to_column("ASV"), 
            by = "ASV")

# Bayesian estimation of dispersion
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)
plotDispEsts(dds)


ggplot(sigtab, aes(x=log2FoldChange, y=Class, color=Division)) + 
 geom_jitter(size=2, width = 0.2) +  
 scale_color_manual(values = distinct_palette(n = NA, pal = "brewerPlus")  ) 



# MA plots
# "Log fold-change vs. mean shows how well the homoskedasticity assumption holds,
# and identifies unusual OTUs where fold-change is at a limit."
plotMA(res)


```
# heatmap
```{r}

select <- sigtab$ASV
nt <- normTransform(dds) # defaults to log2(x+1)
log2.norm.counts <- assay(nt)[select, ]

df <-  colData(dds) %>%  as.data.frame() %>%  dplyr::select(well_id.y)
df2 <- log2.norm.counts %>% as.data.frame() %>% 
  rownames_to_column("ASV") %>% 
  left_join(. , tax_table %>% rownames_to_column("ASV"), 
            by = "ASV") %>% 
  column_to_rownames("ASV") %>% 
  dplyr::select("Division") %>% 
  mutate(Division = as.factor(Division))
  
  
 pheatmap::pheatmap(log2.norm.counts %>% t(), 
                    annotation_col=df2,
                    annotation_row = df,
                    main="" ,  
                    show_colnames = F,
                    show_rownames = T) +
  scale_color_manual(values = distinct_palette(n = NA, pal = "brewerPlus")) + 
  scale_fill_manual(values = distinct_palette(n = NA, pal = "brewerPlus"))
```

